The evolution of gene expression in the term placenta of viviparous mammals

### 2016-04-07 manuscript/manuscript.Rnw
clean up typos and figure references

  \parttitle{Results} We measured the expression profiles of
  \Sexpr{star.logs[,length(unique(file))]} term placenta of
  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
  found lineage-specific differential expression of prolactin,
  relaxin, and galectins as well as increased expression of the
  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
  \textit{PAPPA} in humans. We also identified genes which are
  expressed in all placentas, including genes with immune-modulating
  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
  \textit{LUM}, and \textit{LGALS1}).

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
the match with the highest bitscore was chosen as the annotation;
transcripts where the highest scoring match had a bitscore below 40
were left unannotated.
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
    temp.table <- 
        lineage.table[analysis.name==analysis &
                      p.value <= 0.05][order(p.value),
                                       list(human_name,p.value,fdr,fc),]
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
        if (orderbyfc) {
            display.table <- temp.table[order(-abs(`Fold Change`)),]
        }
        display.table <- display.table[1:members,]

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB4} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of annexin complexes were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family \cite{Green.ea2000:Pregnancy-associatedbovineandovine} 
show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\text{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

@@

\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                                     Species),],
             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(),
              axis.text.y = element_text(face="italic"),
              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
  studied with housekeeping
  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  removed.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

common.plot.options <- 
    list(geom_tile(aes(fill = fpkm), colour = "white"),
         scale_fill_gradient(low = grad.def[["low"]],
                             high = grad.def[["high"]],
                             space = grad.def[["space"]],
                             na.value = grad.def[["na.value"]],
                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
         scale_x_discrete("", expand = c(0, 0)),
         scale_y_discrete("", expand = c(0, 0)),
         theme_grey(base_size = 9),
         theme(legend.position = "top",
               axis.ticks = element_blank(),
               panel.background = element_blank(),
               axis.text.y = element_text(face="italic"),
               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
         facet_grid(type~.,scale="free_y",space="free_y")
         )
print(ggplot(combined.long, aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\caption{Expression in $\text{log}_{10}(\text{FPKM})$ of 1:1 \human
  orthologs of Hormone Genes, Galectins, genes implicated in
  pre-eclampsia (PE), and \textit{IGF}-related genes}
\label{fig:hormone_genes}
\end{figure}

>>>>

  \parttitle{Results} We measured the expression profiles of
  \Sexpr{star.logs[,length(unique(file))]} term placentas of
  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
  found lineage-specific differential expression of prolactin,
  relaxin, and galectins as well as increased expression of the
  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
  \textit{PAPPA} in humans. We also identified genes which are
  expressed in all placentas, including genes with immune-modulating
  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
  \textit{LUM}, and \textit{LGALS1}).

@@

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
the match with the highest bitscore was chosen as the annotation;
transcripts where the highest scoring match had a bitscore below 40
were left unannotated.
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
%

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\caption{Species tree for organisms studied showing the relationship
  between species. Silhouettes of species were obtained from
  \url{http://phylopic.org}. }
\label{fig:species_tree}
\end{figure}

@@

placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
    temp.table <- 
        lineage.table[analysis.name==analysis &
                      p.value <= 0.05 &
                      fdr <= fdr.cutoff][order(p.value),
                                       list(human_name,p.value,fdr,fc),]
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
        if (orderbyfc) {
            display.table <- temp.table[order(-abs(`Fold Change`)),]
        }
        display.table <- display.table[1:members,]

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB4} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of annexin complexes were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family \cite{Green.ea2000:Pregnancy-associatedbovineandovine} 
show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $\text{FPKM} \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\text{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B} %
(\Sexpr{gene.expression("bos taurus","HSD3B")}).

@@

\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
cptl <-
    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
                                            Species),]
cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
                                          capfirst(names(species.names))))]
print(ggplot(cptl,
             aes(y=human_name,x=Species))
      + geom_raster(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low = grad.def[["low"]],
                            high = grad.def[["high"]],
                            space = grad.def[["space"]],
                            na.value = grad.def[["na.value"]],
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(),
              axis.text.y = element_text(face="italic"),
              axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic"))
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
  studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. Supplemental
  \cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different FPKM and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

common.plot.options <- 
    list(geom_tile(aes(fill = fpkm), colour = "white"),
         scale_fill_gradient(low = grad.def[["low"]],
                             high = grad.def[["high"]],
                             space = grad.def[["space"]],
                             na.value = grad.def[["na.value"]],
                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
         scale_x_discrete("", expand = c(0, 0)),
         scale_y_discrete("", expand = c(0, 0)),
         theme_grey(base_size = 9),
         theme(legend.position = "top",
               axis.ticks = element_blank(),
               panel.background = element_blank(),
               axis.text.y = element_text(face="italic"),
               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")),
         facet_grid(type~.,scale="free_y",space="free_y")
         )
print(ggplot(combined.long, aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\caption{Expression in $\text{log}_{10}(\text{FPKM})$ of 1:1 \human
  orthologs of Hormone Genes, Galectins, genes implicated in
  pre-eclampsia (PE), and \textit{IGF}-related genes. Empty locations
  indicate no detectable expression and/or no 1:1 \human ortholog in
  that species.}
\label{fig:hormone_genes}
\end{figure}
